A Full Accounting of Landcover Map Error and Bias and Their Impacts on Assessments of Global Change

### 2015-09-24 paper/main.tex
Updated through results: changed terminology for absolute errors, updated all figures and Rmds creating them accordingly.

\title{The Impact of Bias in Landcover Maps}
%\title{Our Understanding of Global Change is Built on a Shaky Foundation}

@@

Just as a tall edifice cannot be built on a shaky foundation, global change science needs to be based on sound landcover data. There is thus an urgent need to more precisely quantify landcover map errors and how these vary over large regions, particularly for the regions where landcover is changing most rapidly yet is most poorly known.  We address this need in this study, using a unique, high accuracy agricultural landcover map for the entire country of South Africa to quantify the errors in several latest generation landcover maps that are widely used in global change studies.  We use this information to examine how i) landcover properties and related classification schemes influence error, ii) how these errors change with aggregation scale, with the specific goal of determining ``safe'' scales for drawing area-based inferences, and iii) how these errors propagate through several different forms of downstream analyses that broadly represent the global change research focus areas, including biogeochemical and land use change studies, food security assessments, land surface hydrology and climatology, and human geography.

@@

We compared the reference map with similar maps derived from four existing landcover datasets. We obtained South Africa's 30 m resolution National Landcover map (SA-LC) for 2009 \cite{sanbi_national_2009}, the 500 m resolution MODIS Landcover for 2011 \cite{land_processes_distributed_active_archive_center_lp_daac_modis_2011, friedl_modis_2010}, the 300 m resolution GlobCover 2009 \cite{arino_global_2012}, and the new 1 km Geo-wiki hybrid-fusion cropland map for Africa \cite{fritz_mapping_2015}. We chose these particular datasets because they are nearly contemporaneous with our reference data, and represent the major types of landcover products used by researchers: SA-LC typifies the higher resolution, Landsat-derived maps that are developed individually for many countries \cite[e.g.][]{fry_completion_2009}, MODIS and GlobCover are widely used global-scale products \cite{gross_monitoring_2013,shackelford_conservation_2015}, while Geo-Wiki incorporates the first three datasets and represents the current state-of-the-art in landcover mapping. We extracted the cropland classes from the first three datasets and converted these to 1 km resolution percent cropland estimates (hereafter simply the ``cropland maps''), resulting in 4 maps to compare to our reference map.

\vspace{-0.5 cm}
\section{Quantifying Error and Bias}
We first quantified the errors in cropland area estimates based on the pixel-wise differences between the reference map and each of the four cropland maps. We calculated these errors for five different scales of aggregation, from the original 1 km up to 100 km, in order to calculate bias--the average pixel-wise error--and how it varies with scale. Next, we assessed the degree to which the pattern of agricultural landscapes, expressed in terms of the mean cropland cover, impacts bias.

We then investigated how these landcover map biases propagate through data that are built upon them. We began by examining the biases in three different kinds of landcover-based data that we consider to be ``first-order'', in that they are derived by mapping a variable of interest to a landcover type(s). The first of these was to assess potential biases in Tier-1 carbon accounting maps that we constructed from the cropland maps, following \cite{ruesch_new_2008}. The second was an assessment of gridded maize yields derived by disaggregating district-scale agricultural census data for both maize yield and harvested area \cite[following][]{monfreda_farming_2008,ramankutty_farming_2008}. We used these same data to calculate gridded maize production estimates, which we used to conduct the third first order analysis.

Finally, we conducted two second-order analyses, in which a process model draws on the cover types' values to calculate an output value. For the first of these, we examined how cropland map biases influence 25 km resolution monthly evapotranspiration calculated by the Variable Infiltration Capacity \cite{liang_simple_1994} land surface hydrology model. For this example, we used the cropland maps to adjust landcover-specific leaf area index (LAI) values that VIC uses to partition water vapor fluxes into their evaporative and transpirative components. In the second example, we examined how these errors can impact the parameterization of an agent-based food security model \cite{chen_dependency_2013}. Spatially-explicit, agent-based models are frequently employed in land change science, and require an initialization step to assign landscape resources to model agents  \cite[e.g.][]{manson_agent-based_2007,evans_multi-scale_2004,kelley_relative_2011}. In this case, we used cropland maps to provide the model the location and abundance of cropland, which is used to  allocate an initial share of cropland to each farm household (agent) in the model.

\vspace{-0.5 cm}
\section{Bias}
\subsection{Landcover bias}

@@

We then aggregated the reference and each cropland map to 5, 10, 25, 50, and 100 km resolutions, and subtracted the four landcover maps from the reference map at each scale of aggregation to assess patterns of differences (Fig. 1). Negative pixels here represent overestimation error, while positive values indicate underestimates.

\vspace{-0.25 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/bias_map.pdf}}
\caption{Differences in percent cropland estimates between the reference map and each of the four landcover maps. Rows indicate the landcover map being assessed (by subtraction from the reference map), while columns refer to resolution of aggregation. White indicates areas with no data where communal farmlands or plantation forests were removed.}\label{afoto}
\end{figure}

The most pronounced differences were in the MODIS and GlobCover maps, which both underestimated cropland extent by 10-75\% in the center of the country (blue areas in Fig. 1, and the dominant production region), and overestimated along the eastern to northern margins (red areas in Fig. 1). The differences between the reference and these two maps at 1 km resolution were respectively 21\% (MODIS) and 34\% (GlobCover), on average. This mean pixel-wise error expresses the amount of \emph{bias} in a map, that is its tendency to err in a particular direction. For MODIS, bias drops to 8\% at 50 km of aggregation, whereas GlobCover bias is 24\% at 100 km (SI Appendix, Fig. S1).

The SA-LC map uniformly overestimated cropland throughout the country (Fig. 1), but was less biased, ranging from -8\% at 1 km to -6\% at 100 km  (Fig. S1). The GeoWiki map has a strongly heterogenous pattern of differences (Fig. 1) and the smallest magnitude of bias, which changed between slight tendencies to underestimate (5\% at 1 km) and overestimate (-2 \% at 100 km,  Fig. S1).

\subsection{The role of landscape characteristics in shaping bias}
To investigate the degree to which landscape features influence landcover map bias, we extracted all pixels in agricultural areas ($>$0\% cropland) of the 1 km reference map using the boundaries of 354 magisterial districts (South Africa's finest administrative unit, which average 3,445 km$^2$ in size; SI Appendix, Fig. S3), and calculated the district-wise mean for these pixels to provide a measure of cropland density, and information on the degree of mixing between cropland and other land covers. We then extracted the cropland map errors for the same pixels, and calculated the absolute value of bias for each district. Absolute bias describes the magnitude of bias but not its direction, and is more informative about the likelihood of the map being biased at any given point in the landscape than actual bias, as positive and negative biases can cancel each other out within small areas.

A generalized additive model fit to district-level absolute bias (log-transformed) shows that absolute bias peaks at 50\% cropland cover for all but the GlobCover map (which continued to increase with cropland cover), and is lowest when the landscape is dominated either by cropland or other cover types (Fig. 2). In other words, bias is highest when cropland cover is mixed evenly with other cover types.

%\vspace{-1 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.4\textwidth]{figures/biases_md_lnorm_gam_mu0.pdf}}
\caption{The relationship between the mean absolute bias in cropland maps and cropland cover within agricultural areas (reference map pixels having $>$0.5\% cropland), averaged within the boundaries of magisterial districts (n = 345), as fit with a generalized additive model. Prediction curves are color-coded to the different cropland maps, with the solid line indicating  predicted absolute bias, and the lighter shading the standard error of the coefficients. Models were also fit to the mean absolute biases across all four maps (black curve), and from all maps exlcuding GlobCover (grey surve).}
\label{afoto}
\end{figure}

\subsection{The impact of bias on calculating carbon stocks}
We used the data of \cite{ruesch_new_2008} to calculate carbon densities for African forests, shrublands, croplands, grasslands, and sparse habitats (semi-arid grasslands and low shrublands), assigning cropland carbon values to map cells in proportion to their cropland cover. For the non-cropland proportions, we assigned the carbon value from each of the other types, such that we created four different carbon maps for each landcover map at each aggregation scale (Fig. S4), which allowed us to test how carbon estimates vary as a function of i) cropland map bias and ii) the characteristics of adjacent cover types.

@@

The spatial patterns of errors in carbon estimates (Fig. S4) reflect those of cropland biases (Fig. 1). Where cropland was underestimated and the surrounding cover type was of higher carbon density than cropland, carbon density was overestimated. For lower density cover (grassland and sparse vegetation), carbon stocks were underestimated, but by small magnitudes. These tendencies were reflected in each map's biases, calculated over agricultural areas (Fig. 3). For example, MODIS and GlobCover bias was $\sim$-50\% (overestimation) at 1 km resolution when forest was the cropland-adjacent cover (stars in semi-transparent green and gold bars, Fig. 3; Table S2). For sparse vegetation (open circles in Fig. 3), MODIS bias was 3\% at all scales, whereas maps that overestimated cropland (e.g. SA-LC, semi-transparent red) overestimated carbon density for this cover type, because cropland has a higher carbon density \cite{ruesch_new_2008}. Overall, GeoWiki had the lowest bias, for all cover types and all resolutions. It worst bias was a tendency to overestimate by 12\% at 1 km when forest was adjacent, but at coarser scales this bias reduced to just a few percent (Fig. 3, Table S2). All maps' biases are within +/-10\% bias after aggregation to 25 km.

%\vspace{-0.9 cm}
\begin{figure}[!ht]
\centerline{\includegraphics[width=.4\textwidth]{figures/carbon_veg_scale3.pdf}}
\caption{The biases in carbon density estimates based on cropland maps, expressed as a percent difference relative to the reference map. The values of actual bias (incorporating bias direction and magnitude) and absolute bias (magnitude only) are presented in semi-transparent and solid colors, respectively, with colors denoting the specific cropland map and symbols indicating which cover type was used to calculate cropland adjacent carbon density. The bar represents the mean biases calculated across each of the 5 cover types. Shrubland and grassland bias values are near zero and not shown for display clarity.}
\label{afoto}
\end{figure}

Absolute bias in carbon maps (solid colored bars in Fig. 3) generally followed the same patterns, but with somewhat higher magnitudes and a few important differences. The most notable is that Geo-Wiki had fairly large absolute bias at 1 km, averaging 27\% across across cover types (line in solid blue bar, Fig. 3, Table S2), which is close to the 36-37\% for GlobCover and MODIS. SA-LC had the lowest absolute bias across scales, averaging (across cover types) from 14\% at 1 km to 3\% at 100 km. (Fig. 3, Table S2). The increase in Geo-Wiki's absolute bias relative to SA-LC's can be attributed to the highly heterogeneous nature of its cropland errors (Fig. 1), which varied between positive and negative over much shorter distances than the other three cropland maps.

\subsection{Bias in harvested areas, yield, and production estimates}
The disaggregated yield and harvested area maps of \cite{monfreda_farming_2008} are built upon cropland fraction maps where the total area is adjusted to match survey-derived agricultural area statistics reported for administrative districts \cite[provinces, in South Africa's case][]{ramankutty_farming_2008}. To be consistent with this methodology, we adjusted our cropland maps using the same procedure, but we used the reference map to calculate total cropland area for each of South Africa's nine provinces, and then updated the pixel-wise cropland percentages in the four cropland maps so that the province-wise sums matched the reference areas \cite[][and see SI]{ramankutty_farming_2008}. Despite this statistical constraint, the updated cropland maps still had large spatial errors that were similar in pattern (Fig. S5) to those in the unadjusted maps (Fig. 1), and we evaluated how these residuals affected gridded estimates of the yield and production of maize, South Africa's largest crop \cite{estes_comparing_2013}. To create these maps, we followed \cite{monfreda_farming_2008} by disaggregating district-level (n = 354, mean area = 3,445 km$^2$) agricultural census data \cite{statistics_south_africa_commercial_2007} for maize yield and harvested area, aggregated each set of maps, and multiplied the two to calculate production at each scale.

@@

%\vspace{-1 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.4\textwidth]{figures/yield_bias_scale.pdf}}
\caption{The biases in disaggregated maize yield and production estimates. The values of actual bias (incorporating bias direction and magnitude) and absolute bias (magnitude only) are presented in semi-transparent and solid colors, respectively, with colors denoting the specific cropland map and symbols indicating whether the bias is related to one of two yield aggregation methods and associated crop production estimates.}
\label{afoto}
\end{figure}

Production errors were completely unbiased (Fig. 5). The statistical constraints on harvested and cropland areas resulted in the canceling out of spatial errors in production estimates, which is evident in the checkerboard-like pattern in maps of production biases (Figure S7). However, this pattern of bias means that absolute bias in production estimates were large, between 40 to 55\% for GeoWiki, MODIS, and GlobCover at 1 km, and remained generally high (10-28\%) even up to 50 km of aggregation (Fig. 5). SA-LC production biases were lowest across all spatial scales (20\% at 1 km to 2\% in 100 km).

Absolute mean biases in yield were also substantial, and generally 10-15\% larger than production biases across all spatial scales, except for GlobCover where absolute production biases exceeded yield bias at 5-100 km of aggregation.

\subsection{Bias in evapotranspiration estimates}
Compared to carbon and crop related examples, actual and absolute biases in evapotranspiration calculated using the VIC model were small and averaged to less than +/-1\%. However, there are several hotspots of discrepancy evident in the error maps (Fig. 6). The most pronounced of these are the 5-15\% overestimates in the center of the country that resulted when VIC was initialized with MODIS and GlobCover, while overestimates along the southern and western coasts reached 25\%. These locations correspond primarily to the margins of major crop production regions--in the center is the westernmost boundary of the summer rainfall growing region, where maize is the primary crop, while the hotspot on the west coast marks the western boundary of the wheat-dominated winter rainfall region \cite{hardy_rainfed_2011}. These two boundaries respectively correspond to the 400 and 200 mm isohyets of growing season rainfall.

SA-LC and GeoWiki also produced biased ET estimates along the southern and western coasts, but here the tendency was to underestimate ET, while biases in the center of the country were either negligible to absent.  All but MODIS underestimated ET by 5-15\% in the northern tip of the country.

\FloatBarrier
\begin{figure}[ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/et_bias_map.pdf}}
\caption{Differences in annual mean evapotranspiration estimates from 29-year runs of the VIC land surface hydrology model when initialized with LAI response curves derived from the reference map, versus those from the four cropland maps.}\label{afoto}
\end{figure}

\subsection{Initialization biases in agent-based models}
We used an agent-based model (ABM) of food security that represents the interactions between hundreds of individual farming households over multiple seasons \cite{chen_dependency_2013}. Like many spatial ABMs, the model is computationally intensive, and thus run over smaller geographic domains (e.g. districts, rather than an entire country) and at higher spatial resolutions (10s to 100s of meters) that are needed to represent the different land units of single farmers. To match these computational characteristics, we selected four contiguous magisterial districts (ranging from 1,040-1,343 km$^2$, Fig. S8) in the eastern part of the country, having between 28-45\% of their areas devoted to cropland, according to the reference map.

@@

%randomly siting 100 household agents within each district, and allocating the nearest two cropland pixels to each household. The remaining agents are then iteratively assigned unallocated cropland pixels within a 1.5 km radius of existing agents' fields, and this process continues until all agents are assigned cropland, or all available cropland is allocated. This initialization process

@@

\begin{itemize}
  \item First large area quantification of spatial biases
  \item How large those biases are, for one of the most widely spread (spreading landcovers)
  \item Insight into causes of bias, and thus some understanding of where biases are likely to be greater or smaller
  \item How much progress made in reducing it
  \item Class type and bias
\end{itemize}

Some points to make thrown out now because out of time. Others will be added. Please add any important implications/considerations you see from the results.

Some points from my notes last year: 
\begin{itemize}
  \item Bias as a function of scale
    \begin{enumerate}
      \item At 1 km resolution all landcover products are still fairly biased.  
       \item Bias drops to acceptable levels quickly for geowiki--at 5X5 km, mean bias is just 1\% (overestimated). The absolute bias for this dataset is 10\% or lower from 10X10 km resolution and coarser.  
       \item The SA dataset's bias is fairly consistent but low across all levels of aggregation, amounting to no more than an 8\% overestimate of cropland with absolute bias of similar magnitude. 
       \item MODIS and GlobCover biases (mostly of underestimation) do not dissipate until the higher levels of aggregation. MODIS's actual bias (under-estimation) falls below 10\% at 20 km resolution, but the absolute bias remains above 10\% until more than a 100-fold aggregation is done ($>$100 km resolution).  For GlobCover, it is still too high.  
    \end{enumerate}
  \item Bias as a function of cropland cover
    \begin{enumerate}
      \item Classification algorithms are thus more error-prone where landcover is mixed/heterogenous.
      \item The exception to this lies in the GlobCover dataset, where bias primarily increases as a function of cropland cover. The reason for this is that GlobCover's cropland classes do not provide for 100\% cropland cover, so aggregation tends to exacerbate underestimates. 
      \item Thus caution is needed when aggregating a mixed pixel class.  
        \begin{itemize} 
          \item An example illustrats this: take 4 1 km pixels, 2 of which are 100\% cropland, 2 of which are other cover types. Imagine a landcover product classifies 3 of these as cropland (2 correct, 1 an error of commission), using a cropland class that is defined as 50\% cropland. Aggregating the actual fraction by a factor of 4 will result in a new 4 km pixel having 50\% cropland, whereas aggregating the landcover product's pixels will give just 38\% cropland, even when factoring in the incorrect classification.)
          \end{itemize}
    \end{enumerate}
  \item Bias as a function of method
      \begin{enumerate}
        \item Higher resolution and ensemble-based approaches have less bias.
        \item geowiki represents a fusion of multiple coarse resolution data sources that has undergone extensive validation using a crowdsourcing approach
        \item the SALC dataset is based on 30 m landsat data, but incorporates a range of ancillary data and expert judgement
        \item MODIS and GlobCover data are effectively single source/single algorithm.
        \item \textbf{Newer points begin here}

>>>>

\title{A Full Accounting of Landcover Map Error and Bias and Their Impacts on Assessments of Global Change}
%\title{Our Understanding of Global Change is Built on a Shaky Foundation}

@@

Just as a tall edifice cannot be built on a shaky foundation, global change science needs to be based on sound landcover data. There is thus an urgent need to more precisely quantify landcover map errors and how these vary over large regions, particularly for the regions where landcover is changing most rapidly yet is most poorly known.  We address this need in this study, using a unique, high accuracy agricultural landcover map for the entire country of South Africa to conduct a spatially comprehensive, bottom-up quantification of error in several latest generation landcover maps that are widely used in global change studies. We use these errors to assess the extent of bias in i) landcover data, ii) how  landcover properties influence this bias, iii) how these biases change with aggregation scale, with the specific goal of determining ``safe'' scales for drawing area-based inferences, and iv) how these biases propagate through several different forms of downstream analyses that broadly represent the global change research focus areas, including biogeochemical and land use change studies, food security assessments, land surface hydrology and climatology, and human geography.

@@

We compared the reference map with similar maps derived from four existing landcover datasets. We obtained South Africa's 30 m resolution National Landcover map (SA-LC) for 2009 \cite{sanbi_national_2009}, the 500 m resolution MODIS Landcover for 2011 \cite{land_processes_distributed_active_archive_center_lp_daac_modis_2011, friedl_modis_2010}, the 300 m resolution GlobCover 2009 \cite{arino_global_2012}, and the new 1 km Geo-wiki hybrid-fusion cropland map for Africa \cite{fritz_mapping_2015}. We chose these particular datasets because they are nearly contemporaneous with our reference data, and represent the major types of landcover products used by researchers: SA-LC typifies the higher resolution, Landsat-derived maps that are developed individually for many countries \cite[e.g.][]{fry_completion_2009}, MODIS and GlobCover are widely used global-scale products \cite{gross_monitoring_2013, shackelford_conservation_2015}, while Geo-Wiki incorporates the first three datasets and represents the current state-of-the-art in landcover mapping. We extracted the cropland classes from the first three datasets and converted these to 1 km resolution percent cropland estimates (hereafter simply the ``cropland maps''), resulting in 4 maps to compare to our reference map.

\vspace{-0.5 cm}
\section{Quantifying Error and Bias}
We first quantified the errors in cropland area estimates based on the pixel-wise differences between the reference map and each of the four cropland maps. We calculated these errors for five different scales of aggregation, from the original 1 km up to 100 km, in order to calculate bias (the mean pixel-wise error) and mean absolute error (which measures the average error magnitude), and how these vary with scale. Next, we assessed the degree to which the average cropland cover in agricultural landscapes, which relates to their pattern, impacts error magnitude.

We then investigated how landcover map error propagates through ``downstream'' analyses that are built on landcover data, and which are typical of those performed in global change assessments. We began by examining the biases and mean absolute errors in three ``first-order'' landcover-based analyses, in which a variable of interest is mapped onto a landcover type(s) using a simple empirical relationship. The first of these was the widely used International Panel on Climate Change's Tier-1 approach for mapping vegetative carbon stocks, as developed by \cite{ruesch_new_2008}. The second was maize yield maps derived by disaggregating district-scale agricultural census data for both maize yield and harvested area \cite[following][]{monfreda_farming_2008,ramankutty_farming_2008}, from which we calculated the third map, gridded maize production estimates. Maps based on these analyses underpin many assessments of crop productivity and production \cite[e.g.][]{foley_solutions_2011,licker_mind_2010}.

Finally, we examined errors resulting from two second-order analyses, in which a process model draws on the cover types' values to calculate an output value. For the first of these, we used the Variable Infiltration Capacity \cite{liang_simple_1994} land surface hydrology model to calculate monthly evapotranspiration, using the reference and cropland maps to adjust landcover-specific leaf area index (LAI) values that VIC uses to partition water vapor fluxes into their evaporative and transpirative components. In the second example, we examined how these map errors impact the parameterization of an agent-based food security model \cite{chen_dependency_2013}. Spatially-explicit, agent-based models are frequently employed in land change science, and require an initialization step to assign landscape resources to model agents  \cite[e.g.][]{manson_agent-based_2007,evans_multi-scale_2004,kelley_relative_2011}. In this case, we used cropland maps to provide the model the location and abundance of cropland, which is used to  allocate an initial share of cropland to each farm household (agent) in the model.

\vspace{-0.5 cm}
\section{Bias}
\subsection{Landcover error and bias}

@@

We then aggregated the reference and each cropland map to 5, 10, 25, 50, and 100 km resolutions, and subtracted the four landcover maps from the reference map at each scale of aggregation to assess error patterns (Fig. 1). Negative pixels here represent overestimation error, while positive values indicate underestimates.

\vspace{-0.25 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/bias_map.pdf}}
\caption{Errors in the percent cropland estimates resulting from each of the four cropland maps relative to the reference map at different scale of pixel aggegation. Rows indicate the landcover map being assessed (by subtraction from the reference map), while columns refer to resolution of aggregation. White indicates areas with no data where communal farmlands or plantation forests were removed.}\label{afoto}
\end{figure}

The most pronounced errors were in the MODIS and GlobCover maps, which both underestimated cropland extent by 10-75\% in the center of the country (blue areas in Fig. 1, and the dominant production region), and overestimated along the eastern to northern margins (red areas in Fig. 1). These average of these were 21\% for MODIS and 34\% for GlobCover at 1 km resolution (Fig. S1), meaning that each map had a strong underestimation bias. Both maps's biases decreased with pixel aggregation, with MODIS biases falling to 8\% at 50 km, but GlobCover's was still 24\% at 100 km (SI Appendix, Fig. S1).

The SA-LC map uniformly overestimated cropland throughout the country (Fig. 1), but its overall bias was relatively small, ranging from -8\% at 1 km to -6\% at 100 km  (Fig. S1). The GeoWiki map had a strongly heterogenous pattern of differences (Fig. 1) and the smallest magnitude of bias, which changed between slight tendencies to underestimate (5\% at 1 km) and overestimate (-2 \% at 100 km,  Fig. S1).

\subsection{Landscape characteristics and error magnitude}
To investigate the degree to which landscape features influence landcover map errors, we extracted all pixels in agricultural areas ($>$0\% cropland) of the 1 km reference map using the boundaries of 354 magisterial districts (South Africa's finest administrative unit, which average 3,445 km$^2$ in size; SI Appendix, Fig. S3), and calculated the district-wise mean for these pixels, providing a measure of cropland density that was informative of the degree of mixing between cropland and other land covers at landscape scales. We then extracted the cropland map errors for the same pixels, and calculated their mean absolute errors for each district.

A generalized additive model fit to district-level mean absolute error (log-transformed) shows that error peaks at 50\% cropland cover for all but the GlobCover map (which continued to increase with cropland cover), and is lowest when the landscape is dominated either by cropland or other cover types (Fig. 2). In other words, bias is highest when cropland cover is mixed evenly with other cover types. The reason that GlobCover bias continued to increase with cropland density was because this dataset's dominant class identified for South African cropland was a mixture of 50-70\% cropland and other vegetation types, which resulted in the capping of GlobCover-derived cropland cover estimates at the intermediate densities defined by this class.

%\vspace{-1 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/biases_md_lnorm_gam_mu0.pdf}}
\caption{The relationship between the mean absolute error (error magnitude) in cropland maps and the actual cropland cover within agricultural landscapes (reference map pixels having $>$0.5\% cropland), here defined by the boundaries of magisterial districts (n = 345), as fit with a generalized additive model. Prediction curves are color-coded to the different cropland maps, with the solid line indicating predicted absolute bias, and the lighter shading the standard error of the coefficients.}
\label{afoto}
\end{figure}

\subsection{Map error and Tier-1 carbon estimates}
Using the methods provided by \cite{ruesch_new_2008}, we calculated average carbon densities for African forests, secondary forests, shrublands, croplands, grasslands, and sparse habitats (semi-arid grasslands and low shrublands), and assigned cropland carbon values to map cells in proportion to their cropland cover. For the non-cropland proportions, we assigned the carbon value from each of the other types, creating five different carbon maps for each landcover map at each aggregation scale (Fig. S4), which allowed us to test how carbon estimates vary as a function of i) cropland map bias and ii) the characteristics of adjacent cover types.

@@

The spatial patterns of errors in carbon estimates (Fig. S4) reflect those of cropland biases (Fig. 1). Where cropland was underestimated and the surrounding cover type was of higher carbon density than cropland, carbon density was overestimated. For lower density cover (grassland and sparse vegetation), carbon stocks were underestimated, but by small magnitudes. These tendencies were reflected in each map's biases, as calculated over the cropped areas of the country as jointly defined by the reference and each cropland map (Fig. 3). For example, MODIS and GlobCover bias was $\sim$-50\% (overestimation) at 1 km resolution when forest was the cropland-adjacent cover (stars in semi-transparent green and gold bars, Fig. 3; Table S2). For sparse vegetation (open circles in Fig. 3), MODIS bias was 3\% at all scales, whereas maps that overestimated cropland (e.g. SA-LC, semi-transparent red) overestimated carbon density for this cover type, because cropland has a higher carbon density \cite{ruesch_new_2008}. Overall, GeoWiki had the lowest bias, for all cover types and all resolutions. Its worst bias was a tendency to overestimate by 12\% at 1 km when forest was adjacent, but at coarser scales this bias reduced to just a few percent (Fig. 3, Table S2). All maps' biases are within +/-10\% bias after aggregation to 25 km.

%\vspace{-0.9 cm}
\begin{figure}[!ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/carbon_veg_scale.pdf}}
\caption{Biases and mean absolute errors in carbon densities derived from cropland maps, calculated as percents relative to the reference map. Bias estimates (represented by symbols) fall within the semi-transparent bars, while mean absolute errors are contained in the solid bars. Bar colors are coded to specific cropland map, while the symbols indicate which cover type was used to calculate cropland adjacent carbon density. The bar represents the mean biases calculated across each of the 5 cover types. Shrubland and grassland bias values were near zero, while secondary forest values were close to forest values, and thus these are not shown for display clarity (but see Table S2).}
\label{afoto}
\end{figure}

The mean absolute error in carbon maps (solid colored bars in Fig. 3) generally followed the same patterns, but with higher magnitudes and a few important differences. The most notable is that Geo-Wiki, despite relatively low bias, had large mean absolute errors at 1 km, averaging 27\% across across cover types (line in solid blue bar, Fig. 3, Table S2), which is close to the 36-37\% for GlobCover and MODIS. SA-LC had the lowest mean absolute error across scales, averaging (across cover types) 14\% at 1 km to 3\% at 100 km. (Fig. 3, Table S2). The increase in Geo-Wiki's absolute bias relative to SA-LC's can be attributed to the highly heterogeneous nature of its cropland errors (Fig. 1), which traded between between positive and negative errors of high magnitude over shorter distances than the other three cropland maps.

\subsection{Map error and gridded agricultural data}
The disaggregated yield and harvested area maps of \cite{monfreda_farming_2008} are built upon cropland fraction maps where the total area is adjusted to match survey-derived cropland area statistics reported for administrative districts \cite[provinces, in South Africa's case][]{ramankutty_farming_2008}. To be consistent with this methodology, we first adjusted our cropland maps according to this procedure, using the reference map to calculate total cropland area for each of South Africa's nine provinces, then updating the pixel-wise cropland percentages in the four cropland maps so that the province-wise sums matched the reference areas \cite[][and see SI]{ramankutty_farming_2008}. Despite this statistical constraint, the updated cropland maps still had substantial errors that were similar in pattern (Fig. S5) to those in the unadjusted maps (Fig. 1), and we evaluated how these residuals affected gridded estimates of the yield and production of maize, South Africa's largest crop \cite{estes_comparing_2013}. To create these maps, we followed \cite{monfreda_farming_2008} by disaggregating district-level (n = 354, mean area = 3,445 km$^2$) agricultural census data \cite{statistics_south_africa_commercial_2007} for maize (South Africa's largest crop by area, \cite{estes_comparing_2013}) yield and harvested area, aggregated each set of maps, and multiplied the two to calculate production at each scale.

@@

%\vspace{-1 cm}
\begin{figure}[!ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/yield_bias_scale.pdf}}
\caption{Biases and mean absolute errors in disaggregated maize yield and production estimates. Bias estimates (represented by symbols) fall within the semi-transparent bars, while mean absolute errors are contained in the solid bars. Bar colors are coded to specific cropland map, while the symbols indicate which the values for the different variables (production and yield).}
\label{afoto}
\end{figure}

Production errors were completely unbiased (Fig. 5). The statistical constraints on harvested and cropland areas resulted in the canceling out of spatial errors in production estimates, which is evident in the checkerboard-like pattern in maps of production biases (Figure S7). However, this reduction in bias comes at the cost of higher error magnitude, as the mean absolute error in production estimates were large, between 40 to 55\% for GeoWiki, MODIS, and GlobCover at 1 km, and remained generally high (10-28\%) even up to 50 km of aggregation (Fig. 5). SA-LC production biases were lowest across all spatial scales (20\% at 1 km, dropping linearly to 2\% by 100 km).

Absolute mean errors in yield were also substantial, and generally 10-15\% larger than production biases across all aggregation scales, except for GlobCover where absolute production biases exceeded yield bias at 5-100 km of aggregation.

\subsection{Map error and evapotranspiration}
Compared to carbon and crop related examples, bias and mean absolute errors in evapotranspiration (ET) calculated using the VIC model were small and averaged to less than +/-1\%. However, there were several hotspots of discrepancy evident in the error maps (Fig. 6). The most pronounced of these are the 5-15\% overestimates in the center resulting from VIC when initialized with MODIS and GlobCover, while overestimates along the southern and western coasts reached 25\%. These locations correspond primarily to the margins of major crop production regions--in the center is the westernmost boundary of the summer rainfall growing region, marked approximately by the 400 mm isohyet, where maize is the primary crop. The west coast hotspot falls at the western edge of the wheat-dominated winter rainfall region \cite{hardy_rainfed_2011}, where growing season rainfall is approximately 200 mm.

SA-LC and GeoWiki also resulted in ET errors estimates along the southern and western coasts, but here the tendency was to underestimate ET, while biases in the center of the country were either negligible to absent.  All but MODIS underestimated ET by 5-15\% in the northern tip of the country.

\subsection{Initialization errors in agent-based models}
We used an agent-based model (ABM) of food security that represents the interactions between hundreds of individual farming households over multiple seasons \cite{chen_dependency_2013}. Like many spatial ABMs, the model is computationally intensive, and thus run over smaller geographic domains (e.g. districts, rather than an entire country) and at higher spatial resolutions (10s to 100s of meters) that are needed to represent the different land units of single farmers. To match these computational characteristics, we selected four contiguous magisterial districts (ranging from 1,040-1,343 km$^2$, Fig. S8) in the eastern part of the country, having between 28-45\% of their areas devoted to cropland, according to the reference map.

@@

%randomly siting 100 household agents within each district, and allocating the nearest two cropland pixels to each household. The remaining agents are then iteratively assigned unallocated cropland pixels within a 1.5 km radius of existing agents' fields, and this process continues until all agents are assigned cropland, or all available cropland is allocated. This initialization process 
%\FloatBarrier
\begin{figure}[!ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/et_bias_map.pdf}}
\caption{Differences in annual mean evapotranspiration estimates from 29-year runs of the VIC land surface hydrology model when initialized with LAI response curves derived from the reference map, versus those from the four cropland maps.}\label{afoto}
\end{figure}

@@

Main points:

1. Previous studies have quantified disagreement over large areas, where

What we found, significance of study 
\begin{itemize}
  \item First large area quantification of spatial biases
  \item How large those biases are, for one of the most widely spread (spreading landcovers)
  \item Insight into causes of bias, and thus some understanding of where biases are likely to be greater or smaller
  \item How much progress made in reducing it
  \item Class type and bias
\end{itemize}

\begin{itemize}
  \item Bias decreases as function of scale
  \item General bias patterns, appropriate use of landcover products, which landcover products
  \item Appropriate scales of inference, by type of product - 
  \item Aggregation improves results for landcover, generally, 
  \item Sensor resolution, statistical resolution, and merging products have high value
   \begin{itemize}
     \item But don't remove spatial bias - absolute bias matters. Statistical constraints seems to just compress spatial biases to higher rates of turnover. Geo-wiki
     \item But of course these types of data are then dependent on how accurate the statistical data are defining the constraint (cite emperor has no data)
    \end{itemize}
   \item Mixed landscapes increase the chances of omission and commission errors by increasing the number of cover classes, or because such landscapes are less spectrally distinct \cite{estes_diylandcover:_2015}
   \item caveats: Only single country in South Africa. More commercial farming than many other countries, but results are still instructive. Analysis of error as function of landscape type suggests that areas where cropland is more mixed with natural vegetation have higher errors. These sorts of landscapes quite common in smallholder-dominated systems, thus suggests that biases may be even higher elsewhere on the continent. 
\end{itemize}

Implications for understanding global change and policy: \\

Increasing awareness of need to have spatial assessments in global change analyses. Do things such as identify areas where yield gaps are high, or how much carbon or biodiversity will be lost to changes in land use, in order to try prioritize development \cite{searchinger_high_2015,newbold_global_2015}, or to understand coupled human-social dynamics, etc. \\

Our finding suggest:
